Cluster Detection from three-dimensional Lensing Mass Map

### 2021-01-08 doc/paper_ms_method_HSCY1/massmap_method.tex

We propose to use the multi-scale NFW atoms, denoted as
$\{\phi_1,...,\phi_N\}$, as the basis atoms of our dictionary.  On the
transverse plane, the NFW atoms follow surface density profiles of the NFW
halos \citep{haloModel-TJ2003-3pt} with comoving scale radii $r_s$ and
truncation radius $c r_s$, where $c$ is the concentration of the NFW halos.  As
the scales of halos are much smaller than the reachable redshift resolution, we
neglect the depth of halo on the line of sight direction and set the NFW atoms'
profiles in the line of sight direction to one-dimensional ($1$-D) Dirac delta
functions as suggested by \citep{LSS-massMap-Glimpse3D-Leonard2014}. The
multi-scale NFW atoms are defined as
\begin{equation}
\begin{split}
\phi_s(\vec{r}_{\theta},z) =&\frac{f}{2 \pi r_s^2 }
F(|\vec{r}_{\theta}|/r_s) \delta_D(z),\\
&  (s=1...N)
\end{split}
\end{equation}
where $\vec{r}_\theta$ is the comoving position in the transverse plane,
\begin{equation}
F(x)=
\begin{cases}
-\frac{\sqrt{c^2-x^2}}{(1-x^2)(1+c)} + \frac{\arccosh
\left(\frac{x^2+c}{x(1+c)}\right)}{(1-x^2)^{3/2}}  & (x<1),\\
\frac{\sqrt{c^2-1}}{3(1+c)} (1+\frac{1}{c+1}) & (x=1),\\
-\frac{\sqrt{c^2-x^2}}{(1-x^2)(1+c)} +
\frac{\arccos\left(\frac{x^2+c}{x(1+c)}\right)}{(x^2-1)^{3/2}} & (1<x\leq c),\\
0& (x>c).
\end{cases}
\end{equation}
$f=1/[\ln (1+c)-c/(1+c)]$. In this work, we fix $c=4$ for the NFW atoms in
different scale frames. Then, we write the basis atoms into the angular
separation coordinates:
\begin{equation}
\begin{split}
\phi_s(\vec{\theta},z) =&\frac{f \chi(z)^2}{2 \pi r_s^2 }
F(|\vec{\theta}|\chi(z)/r_s) \delta_D(z),\\
& (s=1...N)
\end{split}
\end{equation}
where $\chi(z)$ is the comoving distance at the redshift $z$.

@@

The left panel of Figure \ref{fig_corlensKer} shows the lensing kernels for
lenses at five different redshifts as functions of source galaxy redshifts.
The dashed lines are the lensing kernels for spectroscopic redshifts with
neglectable redshift uncertainties. The solid lines are the lensing kernels for
photometric redshifts with redshift uncertainties shown in Figure
\ref{fig_pdfpz}. As shown by the dashed lines in the left panel of Figure
\ref{fig_corlensKer}, the lensing kernels converge to zero for source redshifts
lower than the lens redshift if the uncertainty on the source galaxy
redshift estimation are neglectable. However, as demonstrated by the solid
lines in the same panel, for the source redshifts with large photo-$z$
uncertainties, the lensing kernels do not converge to zero at redshifts lower
than the lens redshift. This is because the galaxies with photo-$z$ estimations
lower than the lens redshifts may be actually located at higher redshifts due
to the photo-$z$ uncertainties.

>>>>

We propose to use the multi-scale NFW atoms, denoted as
$\{\phi_1,...,\phi_N\}$, as the basis atoms of our dictionary.  On the
transverse plane, the NFW atoms follow surface density profiles of the NFW
halos \citep{haloModel-TJ2003-3pt} with comoving scale radii $r_s$ and
concentration $c$. In this paper, we apply a hard truncation on the NFW
profiles at the virial radius: $c r_s$, the influence of different truncation
forms \citep{haloTrunc-Oguri2011} on the mass map reconstruction is left to our
future work.

As the scales of halos are much smaller than the reachable redshift resolution,
we neglect the depth of halo on the line of sight direction and set the NFW
atoms' profiles in the line of sight direction to one-dimensional ($1$-D) Dirac
delta functions as suggested by \citep{LSS-massMap-Glimpse3D-Leonard2014}. The
multi-scale NFW atoms are defined as
\begin{equation}
\begin{split}
\phi_s(\vec{r}_{\theta},z) =&\frac{f}{2 \pi r_s^2 }
F(|\vec{r}_{\theta}|/r_s) \delta_D(z),\\
&  (s=1...N)
\end{split}
\end{equation}
where $\vec{r}_\theta$ is the comoving position in the transverse plane,
\begin{equation}
F(x)=
\begin{cases}
-\frac{\sqrt{c^2-x^2}}{(1-x^2)(1+c)} + \frac{\arccosh
\left(\frac{x^2+c}{x(1+c)}\right)}{(1-x^2)^{3/2}}  & (x<1),\\
\frac{\sqrt{c^2-1}}{3(1+c)} (1+\frac{1}{c+1}) & (x=1),\\
-\frac{\sqrt{c^2-x^2}}{(1-x^2)(1+c)} +
\frac{\arccos\left(\frac{x^2+c}{x(1+c)}\right)}{(x^2-1)^{3/2}} & (1<x\leq c),\\
0& (x>c).
\end{cases}
\end{equation}
$f=1/[\ln (1+c)-c/(1+c)]$. In this work, we fix $c=4$ for the NFW atoms in
different scale frames. Then, we write the basis atoms into the angular
separation coordinates:
\begin{equation}
\begin{split}
\phi_s(\vec{\theta},z) =&\frac{f \chi(z)^2}{2 \pi r_s^2 }
F(|\vec{\theta}|\chi(z)/r_s) \delta_D(z),\\
& (s=1...N)
\end{split}
\end{equation}
where $\chi(z)$ is the comoving distance at the redshift $z$.

@@

The left panel of Figure \ref{fig_corlensKer} shows the lensing kernels for
lenses at five different redshifts as functions of source galaxy redshifts.
The results include both the lensing kernels for spectroscopic redshifts with
neglectable redshift uncertainties and photometric redshifts with redshift
uncertainties shown in Figure \ref{fig_pdfpz}. As shown by the dashed lines in
the left panel of Figure \ref{fig_corlensKer}, the lensing kernels converge to
zero for source redshifts lower than the lens redshift if the uncertainty on
the source galaxy redshift estimation are neglectable. However, as demonstrated
by the solid lines in the same panel, for the source redshifts with large
photo-$z$ uncertainties, the lensing kernels do not converge to zero at
redshifts lower than the lens redshift. This is because the galaxies with
photo-$z$ estimations lower than the lens redshifts may be actually located at
higher redshifts due to the photo-$z$ uncertainties.

### 2021-01-08 doc/paper_ms_method_HSCY1/massmap_sim.tex

We sample halos in a two-dimensional redshift-mass plane. The redshift-mass
plane is evenly divided into eight redshift bins and eight mass bins. We
randomly shift the input halo redshifts and halo masses from the bins' centers
by a small amount. The concentration of the NFW halo is set as a function of
the halo's virial mass ($M_{200}$) and redshift ($z_{h}$) according to
\citet{c-M_Magneticum-Ragagnin2019}
\begin{equation}
c_{h}=6.02\times\left(\frac{M_{200}}{10^{13} M_{\odot}}\right)^{-0.12}
\left(\frac{1.47}{1.+z_h}\right)^{0.16}.
\end{equation}
The weak-lensing shear fields of these NFW halos are simulated according to
\citet{haloModel-TJ2003-3pt}. The shear distortions are applied to one hundred
realizations of galaxy catalogs with the HSC-like shape measurement error and
photo-$z$ uncertainty.

>>>>

We sample halos in a two-dimensional redshift-mass plane. The redshift-mass
plane is evenly divided into eight redshift bins and eight mass bins. We
randomly shift the input halo redshifts and halo masses from the bins' centers
by a small amount. In the simulation, we set the non-linear over-density
($\Delta_{\rm{vir}}$) to $200$, and $M_{200}$ refers to the virial mass. The
concentration of the NFW halo is set to as a function of the halo's virial mass
($M_{200}$) and redshift ($z_{h}$) according to
\citet{c-M_Magneticum-Ragagnin2019}
\begin{equation}
c_{h}=6.02\times\left(\frac{M_{200}}{10^{13} M_{\odot}}\right)^{-0.12}
\left(\frac{1.47}{1.+z_h}\right)^{0.16}.
\end{equation}
The halos are truncated at the virial radius.
The weak-lensing shear fields of these NFW halos are simulated according to
\citet{haloModel-TJ2003-3pt}. The shear distortions are applied to one hundred
realizations of galaxy catalogs with the HSC-like shape measurement error and
photo-$z$ uncertainty.

### 2021-01-08 doc/paper_ms_method_HSCY1/ms.tex

\begin{abstract}
A new method is developed to reconstruct three-dimensional mass maps from
photometric weak-lensing shear measurements. The three-dimensional mass map is
modeled as a summation of the NFW basis atoms, which have two-dimensional
multi-scale NFW surface density profiles on the transverse plane and
one-dimensional Dirac delta functions in the line of sight direction.  The
adaptive lasso algorithm is applied to find a sparse reconstruction of the mass
maps.
We study the performance of cluster detection from the reconstruction mass map
with simulations that applies shear field from isolated halo to HSC-like galaxy
shapes and photometric redshifts.
Our findings are summarized as follows:
1) The lasso solution suffers from a smear of structure in the line of sight
direction even in the absence of shape noise, and in contrast, the adaptive
lasso algorithm efficiently removes the line of sight smear.
2) The algorithm is able to detect halo with minimal mass limits of $10^{14.0}
M_{\odot}/h$, $10^{14.7} M_{\odot}/h$, $10^{15.0} M_{\odot}/h$ for the low
($z<0.3$), median ($0.3\leq z< 0.6$) and high ($0.6\leq z< 0.85$) redshifts,
respectively, with a false detection of 0.022/deg$^2$.
3) The estimated redshift of the halos detected from the reconstructed mass
maps are lower than the true redshift by about $0.03$ for halos at low
redshifts ($z\leq 0.4$). The relative redshift bias is below $0.5\%$ for halos
at $0.4<z\leq 0.85$.
\end{abstract}

@@

The two-dimensional ($2$-D) density map reconstruction which recovers an
integration of projected mass along the line of sight has been well studied
within the community
\citep{massMap-KS1993,WL-massMap-Glimpse2D-Lanusse2016,sparseBaysianMassMap-Price2020}
and applied to large-scale surveys
\citep{HSC1-massMaps,massMapDES-Chang2018,DES-SV-massMap-sparsity}.
Galaxy clusters are detected from the $2$-D mass map
\citep{WL-massMap-clusDet-CFHT-Shan2012,WL-massMap-clusDet-HSC-Miyazaki2018,WL-massMap-clusDet-HSC-Hamana2020};
however, cross matching with some (e.g., optically selected) cluster
catalog is needed to assign the redshift and extract its physical parameter
(such as mass) for each cluster selected from $2$-D mass map.

On the other hand, \citet{WL-clusDet-Hennawi2005} propose to detect clusters
from smoothed shear measurements after convolving tomographic shear
measurements with $3$-D filters that match the tangential shears induced by
multi-scale NFW halos.
Alternatively, we can directly reconstruct $3$-D mass maps taking advantage of
photometric redshift (photo-$z$) information
\citep{LSS-massMap-Wiener-Simon2009,WL-massMap-VanderPlas2011}; however, these
methods either do not have enough spatial resolution to identify clusters, or
they suffer from smears along the line of sight direction, which need to be
overcome for practical searches of clusters in 3D mass maps.

In this paper, we develop a novel method for high-resolution $3$-D mass map
reconstruction that are free from line of sight smears and study the
performance of cluster detection from the reconstructed mass maps.  The $3$-D
mass map is modeled as a summation of the NFW \citep{halo-NFW1997ApJ} basis
atoms, which have $2$-D multi-scale NFW surface density profiles on the
transverse plane and one-dimensional Dirac delta functions in the line of sight
direction. The adaptive lasso algorithm \citep{AdaLASSO-Zou2006} is applied to
find a sparse reconstruction of the mass maps.

>>>>

\begin{abstract}
A new method is developed to reconstruct three-dimensional mass maps from
photometric weak-lensing shear measurements. The three-dimensional mass map is
modeled as a summation of the NFW basis atoms, which have two-dimensional
multi-scale NFW surface density profiles on the transverse plane and
one-dimensional Dirac delta functions in the line of sight direction.  The
adaptive lasso algorithm is applied to find a sparse reconstruction of the mass
maps.
We study the performance of three-dimensional cluster detection from the
reconstructed mass map with simulations that applies shear distortions from
isolated halo to HSC-like galaxy shapes with realistic photometric redshift
uncertainties .
Our findings are summarized as follows:
1) The lasso reconstructed suffers from a smear of structure in the line of
sight direction even in the absence of shape noise, and in contrast, the
adaptive lasso algorithm efficiently removes the line of sight smear.
2) The algorithm is able to detect halo with minimal mass limits of $10^{14.0}
M_{\odot}/h$, $10^{14.7} M_{\odot}/h$, $10^{15.0} M_{\odot}/h$ for the low
($z<0.3$), median ($0.3\leq z< 0.6$) and high ($0.6\leq z< 0.85$) redshifts,
respectively, with a false detection of 0.022/deg$^2$.
3) The estimated redshift of the halos detected from the reconstructed mass
maps are lower than the true redshift by about $0.03$ for halos at low
redshifts ($z\leq 0.4$). The relative redshift bias is below $0.5\%$ for halos
at $0.4<z\leq 0.85$.
\end{abstract}

@@

The two-dimensional ($2$-D) density map reconstruction which recovers an
integration of projected mass along the line of sight has been well studied
within the community
\citep{massMap-KS1993,WL-massMap-Glimpse2D-Lanusse2016,sparseBaysianMassMap-Price2020}
and applied to large-scale surveys
\citep{HSC1-massMaps,massMapDES-Chang2018,DES-SV-massMap-sparsity}.
Cluster detection from $2$-D mass maps has been widely studied and applied to
these large-scale weak-lensing surveys
\citep{WL-massMap-clusDet-CFHT-Shan2012,WL-massMap-clusDet-HSC-Miyazaki2018,WL-massMap-clusDet-HSC-Hamana2020};
however, cross matching with some (e.g., optically selected) cluster
catalog is needed to assign the redshift and extract its physical parameter
(such as mass) for each cluster selected from $2$-D mass map.

On the other hand, we can directly reconstruct $3$-D mass maps taking advantage
of photometric redshift (photo-$z$) information
\citep{LSS-massMap-Wiener-Simon2009,WL-massMap-VanderPlas2011}; however, these
methods either do not have enough spatial resolution to identify clusters, or
they suffer from smears along the line of sight direction, which need to be
overcome for practical searches of clusters in 3D mass maps.  Alternatively,
\citet{WL-clusDet-Hennawi2005} propose to perform a maximum-likelihood
detection of clusters, by convolving tomographic shear measurements with $3$-D
filters that match the tangential shears induced by multi-scale NFW halos,
without fully reconstruct the $3$-D mass maps.

In this paper, we develop a novel method for high-resolution $3$-D mass map
reconstruction that are free from line of sight smears.  The $3$-D
mass map is modeled as a summation of the NFW \citep{halo-NFW1997ApJ} basis
atoms, which have $2$-D multi-scale NFW surface density profiles on the
transverse plane and one-dimensional Dirac delta functions in the line of sight
direction. The adaptive lasso algorithm \citep{AdaLASSO-Zou2006} is applied to
find a sparse reconstruction of the mass maps.

We study the performance of cluster detection from the reconstructed mass maps
using simulations applying shear distortions from isolated halos to HSC-like
galaxy shapes with realistic photo-$z$ uncertainties.

### 2021-01-08 doc/paper_ms_method_HSCY1/sparse_NFW_results.tex

According to the detection rate measured from the simulation, we predict the
number density of detected cluster using the halo mass function of
\citet{haloMass-Tinker2008}. We use HMF \citep{hmf-Murray2013}, an open-source
package, to calculate the halo mass function.  The predicted halo detection
number density for the setup ($\lambda=5$, $1.5\sigma$ detection threshold) is
shown in Figure \ref{fig_detNum}. The expected cluster number density in total
is $0.49$ $\deg^{-2}$, which is $78.4$ clusters for the first year HSC shear
catalog \citep{HSC1-catalog} with a survey area of $\sim 160 \deg^2$.

>>>>

According to the detection rate measured from the simulation, we predict the
number density of detected cluster using the halo mass function of
\citet{haloMass-Tinker2008}. We use HMF \citep{hmf-Murray2013}, an open-source
package, to calculate the halo mass function.  The predicted halo detection
number density for the setup ($\lambda=5$, $1.5\sigma$ detection threshold) is
shown in Figure \ref{fig_detNum}. The expected cluster number density in total
is $0.49$ $\deg^{-2}$, which corresponds to $78.4$ clusters for the first year
HSC shear catalog \citep{HSC1-catalog} with a survey area of $\sim 160 \deg^2$.
The expected number of detections is similar to the number of $2$-D cluster
detections on the first year HSC shape catalog \citep{HSC1-massMap-cluster}.
